Is Vagueness Beneficial for Hearers? \protect \\ Evidence from Experiments

### 2016-09-25 papers/2016-Special-Issue/v-special-issue-2016.tex
pasted kees changes

\documentclass[doc,floatsintext,apacite]{apa6}
\usepackage{graphicx}
\usepackage{multirow}
\usepackage{color}
\usepackage{booktabs}
\usepackage{layout}
\usepackage{apacite}
\usepackage{datetime}
\usepackage{rotating} % comment out for man mode
\title{Is Vagueness Beneficial for Hearers? \protect \\ Evidence from Experiments}
\date{\today}
\authornote{\today ~\currenttime}
\shorttitle{Is Vagueness Beneficial for Hearers?}
\twoauthors{Matt Green}{Kees van Deemter}
\twoaffiliations{Computing Science\\ University of Aberdeen}{Computing Science\\ University of Aberdeen}

\abstract{Much of everyday language is vague, but the causes of this phenomenon are not well understood. Consequently, it is difficult for the designers of a Natural Language Generation (NLG) system to know when and how to let the system generate vague expressions. The present paper is an attempt to find out what benefits vagueness might have for readers.% , by means of a series of carefully controlled psycholinguistic experiments that focus on the use of vagueness in referring expressions. The work is motivated by NLG but has implications for linguistics, philosophical logic and game theory. 
%
The paper reports on a series of experiments that aim to separate the utility of vagueness (as defined by the existence of borderline cases) from the utility of other factors that tend to co-occur with vagueness. We argue that the evidence so far supports a view of vagueness where the benefits that vague terms exert are due to other influences, rather than to vagueness itself. These factors include: low granularity; the use of evaluative words; the avoidance of overtly numerical words; the existence of comparison strategies; and, lastly and more tentatively, a phenomenon that we call range reduction. Although it is possible that other types of vague expressions (i.e., vague words outside referential noun phrases) behave differently, our work suggests that vagueness itself may not increase the utility of an expression. The paper concludes with a brief discussion of the implications of our work for practical NLG.}

@@

In a series of studies of behaviour modification, \citeA<>{Mishra01042011} manipulated the presentation format of information about quantities in the domains of mental acuity, physical strength, and weight loss. In the weight loss study, participants were told that the study was designed to test the validity of a new (actually fictitious) health index, the HHI (Holistic Health Index). They were told that an ideal HHI score lies in the range of 45 to 55. In a longitudinal study, participants submitted their height, weight, hydration level, gender, and age to a computer each week. Participants were told that two algorithms would be used to compute their HHI, and that it was possible that the two algorithms might give different values initially, but would converge over the course of the study to a single value. They were also told that if the two algorithms did give different values, then the true score lay between the two values. In one condition, which the authors called the precise condition, the two algorithms gave the same score. In the other condition, which the authors called the vague condition, one algorithm added 3\% to the score while the other algorithm subtracted 3\% from the score, yielding a range of values whose midpoint was the same as the two values given in the precise condition.

@@

\subsubsection{General Methodology}
The experiments reported in the present paper put the cost reduction hypothesis to the test. 
The question that we are trying to answer is whether vague expressions are processed more easily by readers than crisp ones. 
Like Lipman, we focus on situations where numerical information is used in order to identify a referent. 
Reference, in other words, will be the linguistic task on which we focus, partly because of the interest that this topic has recently drawn from the NLG community.  
In focussing on benefits for the hearer, we will leave aside the question of audience design, leaving this for later research.

@@

% MG added/moved out of expt 1
There is evidence that when the distance grows between two numbers, they become more easily distinguishable from each other: the \emph{numerical distance effect}, which has been shown for comparing the cardinality of two sets of dots \cite{van123} and for processing Arabic numerals and number words \cite{Dehaene199647}. 
We manipulated the number of dots in each array such that some sets of arrays had smaller numerical distances and others had larger numerical distances. 
Where a number was mentioned in the instructions, it was always in the form of an Arabic numeral.

% MG added/moved out of expt 1
There is evidence that when two numbers are presented with the smaller on the left, this left-side presentation facilitates responses indicating the smaller number: the \emph{Spatial-Numerical Association of Response Codes (SNARC)} effect \cite{dehaene1993mental, gevers2006automatic}. 
We controlled which side the smaller number appeared on to avoid systematic influences of this effect.

% MG General methodology moved to here: per-experiment method sections only talk about deviations from this general method
All the experiments shared the following properties: 
Stimuli were created using the language GNU Octave \cite{eaton:2002} and the Psychophysics Toolbox extensions \cite{ptbx1, ptbx2}.
The position of the dots in each array was randomised per-trial.
The order in which trials were presented was randomised per-participant. 
There were 256 trials, presented in 4 blocks of 64 trials each, between which the participant could rest. 
A MacBook Pro laptop computer with a 13 inch screen presented the stimuli to the participants and recorded responses. 
Participants were recruited using email lists at the University of Aberdeen, and paid ten pounds for participating.
All participants self-reported fluency in English, and had normal, or corrected-to-normal vision.
The experiment was conducted in a quiet room. 
Participants were asked to respond as quickly as possible while avoiding errors.
There was a block of practice trials after which participants could ask any questions, following which the experimenter left the room.

@@

\subsection{Introduction}
We used a forced choice task to elicit responses to crisp or vague forms of instructions that required the participant to choose one of two dot arrays by referring to its cardinality.  
The participant was presented with an instruction in the form \emph{Choose the square with \ldots dots}. 
Then a set of two two dot arrays was presented, each in the form of a square containing some number of dots. 
 %Fig \ref{stimuluse1} shows an example stimulus. 
The participant was required to identify the array that corresponded with the instruction, by pressing the appropriate key, as quickly as possible while avoiding errors. 
Response time and accuracy were recorded for analysis.

@@

\subsection{Hypotheses}
(1) A main effect advantage for vagueness: vague instructions impose a lower cognitive load for the comprehender than crisp alternatives.

(2) A main effect advantage of increasing numerical distance: the task will become easier as the numerical distance increases, because the two arrays are then more discriminable.

(3) An interaction between vagueness and numerical distance: i.e., any facilitation for vagueness should be greater at smaller numerical distances than at larger numerical distances.

\subsection{Method}
Twenty participants were recruited, aged between 18 and 45 with a median age of 26.
On each trial a participant was presented with an instruction to choose one of two dot arrays on screen by reference to its cardinality. 
Following a keypress to indicate that the participant had read the instruction there was a central fixation cross for 1000 ms, and a blank screen for 500 ms, followed by the array (without repetition of the referring expression). 
The arrays would stay on screen until the participant responded (there was no timing-out). 
Response time was measured as the latency between the presentation of the arrays, and the keypress identifying the choice: in this way, the response time was separated from time spent reading the instructions, which is important since we are only interested in the former. 
A response was counted as erroneous if the square with the wrong number of dots was chosen (when the instruction contained a number); 
if the square with the larger number of dots was selected (when the instruction was \emph{Choose the square with few dots}); 
or if the square with the smaller number of dots was selected (when the instruction was \emph{Choose the square with many dots}). % KvD Is this addition defining "erroneous" OK? MG Yes, I moved it though
No feedback was given on correct trials, but there was feedback on error trials in the form of the word ``\textsc{wrong!!}'' which flashed on screen.

\subsection{Results}
\subsubsection{Response times}
Response times (RTs) for trials with erroneous responses were discarded, leading to the loss of 354 trials from 5120, representing 6.9\% of the trials. 
The correct response RTs were trimmed at 2.5 standard deviations for each subject, leading to the loss of a further 160 trials, or 3.4\% of the remaining correct responses. 
Means for response times and error rates are given in Fig. (\ref{resultse1}).
A linear mixed model of RT was built, using as independent variables \emph{vagueness} and \emph{numerical distance} and their interaction, with random slopes for \emph{vagueness} and \emph{numerical distance} over participants. 
\emph{Vagueness} was sum coded: \emph{vague} $= -.5$, \emph{crisp} $= .5$; \emph{numerical distance} was Helmert coded, i.e., the variable is ordered and each level is compared against the mean of all the previous levels to that point. % MG a reviewer said he was not familiar with Helmert contrasts
%Level one of this contrast is numerical distance 5 compared with numerical distance 10; level two is the mean of numerical distance 5 and numerical distance 10 versus numerical distance 15; and level three is the mean of numerical distances 5, 10 and 15 versus numerical distance 20. 
All $p$ values were calculated using the R package \emph{lmerTest} \cite{lmerTest}.
Numbering below corresponds to the numbering of the hypotheses.

(1) RTs were faster for vague instructions than for crisp instructions ($\beta=.109$, $se=.022$, $t=4.9$, $p<.001$).

(2) RTs grew faster as numerical distance increased: level one ($\beta=-.116$, $se=.012$, $t=-9.3$, $p<.05$), level 2 ($\beta=-.103$, $se=.009$, $t=-11.1$, $p<.001$) and level three ($\beta=-.082$, $se=.007$, $t=-11.0$, $p<.05$). 
Since discriminability of the arrays is easier for larger numerical distances, discriminability probably underlies this effect.

(3) Numerical distance and vagueness interacted significantly for larger numerical distances when modelling RT: essentially there were diminishing returns for vagueness as numerical distance increased.% KvD Is it worth saying how they interacted? % MG I think this is answered now?
The interactions at the different levels of numerical distance were: level one: ($\beta=.001$, $se=.016$, $t=.03$, $p=.974$); level two ($\beta=-.039$, $se=.009$, $t=-4.460$, $p<.001$); level three ($\beta=-.043$, $se=.006$, $t=-7.0$, $p<.001$). 
In the crisp conditions RTs started out much slower than in the vague conditions, at the smallest numerical distance, but the two conditions converged to very fast times at the largest numerical distance.

@@

(1) The effect of vagueness on error rates approached significance, with the vague instructions leading to fewer errors ($\beta=.307, se=.173, t=1.8, p=.077$).

(2) Error rates decreased as numerical distance increased: level one ($\beta=-.585, se=.092, ,z=-6.4, p<.001$), level 2 ($\beta=-.434, z=-5.3,p<.001$) and level three ($\beta=-.250,z=-4.1, p<.001$).

(3) Error rates were greater in the crisp conditions than the vague conditions when numerical distance was small, and this difference diminished with increasing numerical distance until it reversed at the biggest numerical distance. % MG NEED TO INSERT STATS FOR THIS INTERACTION

@@

\subsection{Discussion}
Responses were faster and more accurate for vague than crisp instructions. 
Response times and error rates improved as the arrays became more easily discriminable. 
The advantage for vague instructions tailed off as the arrays became more easily discriminable.

The cost reduction hypothesis explains the vagueness advantage by claiming that the vague referring expressions place less cognitive load on the comprehender than the crisp referring expressions. 
It explains the diminishing returns for vagueness in more-discriminable stimuli (i.e., the vagueness by numerical distance interaction) by claiming that load is low in both conditions for the easily-discriminable stimuli, and that therefore there is no extra benefit to be had from vagueness in the easily-discriminable stimuli.

@@

The main result from experiment 1 was that responses were faster and more accurate for vague instructions than for crisp instructions.

Contrast, from Experiment 1, an expression from the vague condition: `the square with few dots' with an expression from the crisp condition: `the square with 5 dots'. 
One difference is that `few' is vague (or at least has the potential for vagueness) and `5' is crisp. 
Another difference is that `few' is a verbal quantifier while `5' is a numerical quantifier, in the sense that a number is mentioned explicitly. 
Since these two differences could not be separated in Experiment 1, the vagueness advantage finding is vulnerable to an alternative interpretation, that what we saw as a vagueness advantage was in contrast an advantage for the verbal form of the quantifier. 
In Experiment 2 we created verbal and numeric versions of each of the vague and crisp instructions so that we could compare vague and crisp conditions while taking account of verbal / numeric format.
%In the present experiment 2 we pitted these alternative interpretations against each other in a factorial design.

In Experiment 1, the participants chose one of two squares: therefore the `vague' quantifiers (e.g., `few') uniquely identified one square. 
Recall our definition of vague  -- ``a word is precise if it describes a well-defined set of objects. By contrast, a word is vague if it is not precise''.  
In Experiment 1, it might be said that the quantifiers in the vague conditions did not realise their potential for vagueness. 
This is because there were no borderline cases of the referent that could make the referent set `not well-defined', and perhaps because using definite articles in the instructions implied that only one option was correct.
Using error feedback in experiment 1 could have exacerbated this.
Experiment 2 used three arrays so that the vague quantifiers always had more than one possible referent, and used indefinite articles in the vague instructions to avoid the impression that only one response counted as correct, and was carried out without error feedback.
An indication that the potential for vagueness was realised in Experiment 2 is that the borderline response was chosen fairly often: 16\% of the time.

In Experiment 2, an item was a referring expression instruction followed by a set of dot arrays defined by triple of numbers, representing the number of dots in the left, middle, and right arrays. 
We used four different triples of numbers: (6,15,24); (16,25,34); (26,35,44); (36,45,54). 
Each set of arrays had the following properties: it comprised three arrays (instead of two as in Experiment 1); the array representing the central number was always presented in the middle of the three; there were two flanking arrays where one had fewer dots than the central array and the other had more.

Examples of crisp and vague versions of the numerical and verbal instructions follow: the examples assume the array (6,15,24) and reference to the smaller number of dots, such that 6 was classified as the expected response; 15 was classified as the borderline response; and 24 was classified as the incorrect response. 
In the {\em vague numerical} condition we used \emph{Choose a square with about 10 dots}. None of the squares displayed contained 10 dots. 10 is slightly closer to 6 than to 15, justifying 6 as the best response and 15 as the borderline response.
In the {\em vague verbal} condition we used \emph{Choose a square with few dots}.
In the {\em crisp numerical} condition we used \emph{Choose the square with 6 dots} and one square always did contain the number mentioned.
In the {\em crisp verbal} condition, we used \emph{Choose the square with the fewest dots}.

\subsection{Hypotheses}
(1) A main effect RT advantage for vagueness.

(2) An RT advantage for vagueness both in \emph{numeric} and in \emph{verbal} instructions.

(3) No large main effect of instruction format. %MG If we are right that vagueness was responsible for the vagueness advantage in e 1, that could also have been a number avoidance advantage, then we would not expect that number avoidance, or, instruction format, plays a role (?)

(4) On the basis of Experiment 1, we would expect an effect of triple, with faster responses to more discriminable triples.

(5) Participants should make more borderline case choices for vague than crisp instructions.

@@

(1) The main effect of \emph{vagueness} was to slow responses down, in contrast with Experiment 1, and offering evidence against hypothesis 1 (vague: 2668 ms; crisp: 2450 ms; a difference of 218 ms; $\beta=.06, se=.01, t=4.6, p<.001$).

(2) Vagueness was disadvantageous in both the \emph{numeric} and \emph{verbal} conditions, offering evidence against hypothesis 2. %MG [? give values of a focussed t test]
The disadvantage for vagueness was greater in the numerical than in the verbal conditions, leading to an interaction effect between vagueness and instruction format ($\beta=-0.13, se= 0.02, t= -6.6, p=0.000$).

(3) There was a significant effect of \emph{instruction format} with numerical conditions attracting longer responses than the verbal conditions: consistent with Experiment 1 (numeric: 3284 ms; verbal 1866 ms; a difference of 1418 ms; $\beta=.37, se=.07, t=5.1, p<.001$).

(4) There was a significant main effect of \emph{triple} ($\beta=.06, se=.008, t=7.0, p<.001$) indicating that response times differed in some way according to which triple was presented. 
However further analysis revealed that there was no consistent smooth trend across different triples. 
This effect seems likely to be due to the very fast responses for the smallest triple, which had the largest ratio difference and so may have been particularly discriminable for participants. 
% MG reviews asked for an explanation of the difference for numeric crisp over items. Also reviews mention ratio differences as something that we did not control.

\subsubsection{Borderline responses}
(5) Participant grand mean percentage of borderline selections was $16.6\%$. A generalized linear mixed model \cite{jaeger2008categorical} was fit to the data for selection of the borderline response, with task, vagueness and item as fixed effects, and with random slopes for task and vagueness and item over participants.  
The distribution of responses over the nearest match square, the borderline square, and the furthest match square are given in Fig, \ref{resultse2}.
Participants were significantly more likely to choose the borderline option for vague instructions than for precise instructions (21.9\% vs 11.3\%, $\beta=.79, se=.25, z=3.2, p<.01$). 
Participants were significantly more likely to choose the borderline square when the instruction used the numerical format rather than the verbal format (30.1\% vs 3.0\%, $\beta=3.57, se=.26, z=13.6, p<.001$).

@@

% KvD: I've tentatively added the following:
Having effectively separated the {\bf cost reduction} hypothesis from the {\bf instruction format} hypothesis, it is important to observe that, in Experiment 2, instruction format went hand in hand with {\bf selection algorithm}: 
 %as shown by the Table of instructions \ref{instructionse2}, 
the instructions that used a verbal instruction format allowed a comparison algorithm, whereas the instructions that used a numeric format allowed a matching algorithm. 
Therefore, our results so far permit the interpretation that what made the instructions in the verbal condition fast is not the fact that they were worded verbally, but that they allowed participants to use a comparison algorithm 
(which is known to be faster than matching). % KvD I assume we need to say more about this, citing some appropriate papers to support this idea.

@@

\subsection{Hypotheses}
(1) Vague instructions are easier for the reader than crisp alternatives (main effect of vagueness)

(2) Comparison is easier for the reader than matching (main effect of selection task)

(3) Effects of vagueness are different depending on whether selection is matching or comparison (interaction effect selection x vagueness).

>>>>

\documentclass[doc,floatsintext,apacite]{apa6}
\usepackage{graphicx}
\usepackage{multirow}
\usepackage{color}
\usepackage{booktabs}
\usepackage{layout}
\usepackage{apacite}
\usepackage{datetime}
\usepackage{rotating} % comment out for man mode

\renewcommand{\baselinestretch}{1.7} % added 21 Sept 2016, copied from CRM book

\title{Is Vagueness Beneficial for Hearers? \protect \\ Evidence from Experiments}
\date{\today}
\authornote{\today ~\currenttime}
\shorttitle{Is Vagueness Beneficial for Hearers?}
\twoauthors{Matt Green}{Kees van Deemter}
\twoaffiliations{Computing Science\\ University of Aberdeen}{Computing Science\\ University of Aberdeen}

\abstract{Much of everyday language use is vague, but the causes of this phenomenon are not well understood. Consequently, it is difficult for the designers of a Natural Language Generation (NLG) system to know when to let the system generate vague expressions. The present paper is an attempt to find out what benefits vagueness might have for readers.% , by means of a series of carefully controlled psycholinguistic experiments that focus on the use of vagueness in referring expressions. The work is motivated by NLG but has implications for linguistics, philosophical logic and game theory. 
%
This article reports on a series of experiments that aim to separate the utility of vagueness (as defined by the existence of borderline cases) from the utility of other factors that tend to co-occur with vagueness. We argue that the evidence so far supports a view where the benefits that vague terms exert are due to other influences, rather than to vagueness itself. These factors include: low granularity; the use of evaluative words; the avoidance of overtly numerical words; the existence of comparison strategies; and, lastly and more tentatively, a phenomenon that we call range reduction. The paper concludes with a brief discussion of the implications for practical NLG.}

@@

In a series of studies of behaviour modification, \citeA<>{Mishra01042011} manipulated the presentation format of information about quantities in the domains of mental acuity, physical strength, and weight loss. In the weight loss study, participants were told that the study was designed to test the validity of a new (actually fictitious) health index, the HHI (Holistic Health Index). They were told that an ideal HHI score lies in the range of 45 to 55. In a longitudinal study, participants submitted their weight to a computer each week. Participants were told that two algorithms would be used to compute their HHI, and that the two might give different values initially, in which case the true score lay between the two values. In one condition, which the authors called the precise condition, the two algorithms gave the same score. In the other condition, which the authors called the vague condition, one algorithm added 3\% to the score while the other algorithm subtracted 3\% from the score, yielding a range of values whose midpoint was the same as the two values given in the precise condition.

@@

\subsubsection{General Methodology}\label{method}
The experiments reported in the present paper put the cost reduction hypothesis to the test. 
The question that we are trying to answer is whether vague expressions are processed more easily by readers than crisp ones. 
Like Lipman, we focus on situations where numerical information is used in order to identify a referent. 
Reference, in other words, will be the linguistic task on which we focus, partly because of the interest that this topic has recently drawn from the NLG community.  
In focussing on benefits for the hearer, we will leave aside the question of audience design, leaving this for later research.

@@

% MG General methodology moved to here: per-experiment method sections only talk about deviations from this general method
%All the experiments shared the following properties: 
Stimuli were created using the language GNU Octave \cite{eaton:2002} and the Psychophysics Toolbox extensions \cite{ptbx1, ptbx2}.
The position of the dots was randomised per-trial.
The order in which trials were presented was randomised per-participant. 
There were 256 trials, presented in 4 blocks of 64 each, between which the participant could rest. 
A MacBook Pro laptop computer with a 13 inch screen presented the stimuli to the participants and recorded responses. 
Participants were recruited using email lists at the University of Aberdeen, and paid ten pounds for participating.
All participants self-reported fluency in English, and had normal, or corrected-to-normal vision.
The experiment was conducted in a quiet room. 
Participants were asked to respond as quickly as possible while avoiding errors.
There was a block of practice trials after which participants could ask any questions, following which the experimenter left the room.

% KvD I've moved the following two paragraphs towards the end of the section.
% MG added/moved out of expt 1

When the distance grows between two numbers, they become more easily distinguishable from each other: the \emph{numerical distance effect}, which has been shown for comparing the cardinality of two sets of dots \cite{van123} and for processing Arabic numerals and number words \cite{Dehaene199647}. 
We manipulated the number of dots in each array such that some sets of arrays had smaller numerical distances and others had larger numerical distances. Where a number was mentioned in the instructions, it was always in the form of an Arabic numeral. 
% MG added/moved out of expt 1
When two numbers are presented with the smaller on the left, this left-side presentation facilitates responses indicating the smaller number: the \emph{Spatial-Numerical Association of Response Codes (SNARC)} effect \cite{dehaene1993mental, gevers2006automatic}. 
We controlled which side the smaller number appeared on to avoid systematic influences of this effect.

% KvD added the following from a previous version of the paper:
There is abundant evidence (e.g., \protect\cite{trick1994small}) that very small (i.e., \emph{subitizable}) quantities are recognised and processed by a distinct psychological mechanism that differs from that used to process larger quantities. We performed a pilot experiment in which we were able to confirm this finding in the experimental settings on which we are focussing in this paper \cite{our-earlier-paper}. As one might expect, we found that, when participants were confronted with an array of two boxes, instructions of the form ``Choose the square with $n$ dots" led to consistently lower Response Times than instructions of the form ``Choose the square with many/few dots" when $2 \leq n \leq 5$; the converse was true for $n>5$. We henceforth focussed our studies on non-subitizable numbers, because it is there that vagueness is expected to have benefits.

@@

We used a forced choice task to elicit responses to crisp or vague forms of instructions that required the participant to choose one of two dot arrays by referring to its cardinality. The participant was presented with an instruction in the form \emph{Choose the square with \ldots dots}. Then a set of two dot arrays was presented, each in the form of a square containing some number of dots. 
 %Fig \ref{stimuluse1} shows an example stimulus. 
The participant was required to identify the array that corresponded with the instruction, by pressing the appropriate key, as quickly as possible while avoiding errors. 
Response time and accuracy were recorded for analysis.

@@

\subsection{Hypothesis}
(H1) A main effect advantage for vagueness: vague instructions impose a lower cognitive load for the comprehender than crisp alternatives.
\subsection{Method}
Twenty participants were recruited, aged between 18 and 45 with a median age of 26.
On each trial a participant was presented with an instruction to choose one of two dot arrays on screen by reference to its cardinality. 
Following a keypress to indicate that the participant had read the instruction there was a central fixation cross for 1000 ms, and a blank screen for 500 ms, followed by the array (without repetition of the referring expression). 
The arrays would stay on screen until the participant responded (there was no timing-out). 
Response time was measured as the latency between the presentation of the arrays, and the keypress identifying the choice: in this way, the response time was separated from time spent reading the instructions, which is important since we are only interested in the former. 
A response was counted as erroneous if the square with the wrong number of dots was chosen (when the instruction contained a number); 
if the square with the larger number of dots was selected (when the instruction was \emph{Choose the square with few dots}); 
or if the square with the smaller number of dots was selected (when the instruction was \emph{Choose the square with many dots}). % KvD Is this addition defining "erroneous" OK? MG Yes, I moved it though
No feedback was given on correct trials, but there was feedback on error trials in the form of the word ``\textsc{wrong!!}'' which flashed on screen. Hypothesis H1 was tested looking both at response times and at error rates.

\subsection{Results}

\subsubsection{Response times}
Response times (RTs) for trials with erroneous responses were discarded, leading to the loss of 354 trials from 5120, representing 6.9\% of the trials. 
The correct response RTs were trimmed at 2.5 standard deviations for each subject, leading to the loss of a further 160 trials, or 3.4\% of the remaining correct responses. 
Means for response times and error rates are given in Fig. (\ref{resultse1}).
A linear mixed model of RT was built, using as independent variables \emph{vagueness} and \emph{numerical distance} and their interaction, with random slopes for \emph{vagueness} and \emph{numerical distance} over participants. 
\emph{Vagueness} was sum coded: \emph{vague} $= -.5$, \emph{crisp} $= .5$; \emph{numerical distance} was Helmert coded, i.e., the variable is ordered and each level is compared against the mean of all the previous levels to that point. % MG a reviewer said he was not familiar with Helmert contrasts
%Level one of this contrast is numerical distance 5 compared with numerical distance 10; level two is the mean of numerical distance 5 and numerical distance 10 versus numerical distance 15; and level three is the mean of numerical distances 5, 10 and 15 versus numerical distance 20. 
All $p$ values were calculated using the R package \emph{lmerTest} \cite{lmerTest}.

(H1$_{rt}$) RTs were faster for vague instructions than for crisp instructions ($\beta=.109$, $se=.022$, $t=4.9$, $p<.001$).

@@

(H1$_{er}$) The effect of vagueness on error rates approached significance, with the vague instructions leading to fewer errors ($\beta=.307, se=.173, t=1.8, p=.077$).

@@

As expected, responses were faster and more accurate for vague than crisp instructions. Post hoc, we formulated some further hypotheses:

(H2) A main effect advantage of increasing numerical distance: the task will become easier as the numerical distance increases, because the two arrays are then more discriminable.

(H3) An interaction between vagueness and numerical distance: i.e., any facilitation for vagueness should be greater at smaller numerical distances than at larger numerical distances.

Looking at Response Times, these two post-hoc hypotheses were confirmed:

(H2$_{rt}$) RTs increased with numerical distance level one ($\beta=-.116$, $se=.012$, $t=-9.3$, $p<.05$), level 2 ($\beta=-.103$, $se=.009$, $t=-11.1$, $p<.001$) and level three ($\beta=-.082$, $se=.007$, $t=-11.0$, $p<.05$). Since discriminability of the arrays is easier for larger numerical distances, discriminability probably underlies this effect.

(H3$_{rt}$) Numerical distance and vagueness interacted significantly for larger numerical distances when modelling RT: essentially there were diminishing returns for vagueness as numerical distance increased.% KvD Is it worth saying how they interacted? % MG I think this is answered now? 
The interactions at the different levels of numerical distance were: level one: ($\beta=.001$, $se=.016$, $t=.03$, $p=.974$); level two ($\beta=-.039$, $se=.009$, $t=-4.460$, $p<.001$); level three ($\beta=-.043$, $se=.006$, $t=-7.0$, $p<.001$). 
In the crisp conditions RTs started out much slower than in the vague conditions, at the smallest numerical distance, but the two conditions converged to very fast times at the largest numerical distance.

Looking at Error Rates, the two post-hoc hypotheses were confirmed as well:

(H2$_{er}$) Error rates decreased as numerical distance increased: level one ($\beta=-.585, se=.092, ,z=-6.4, p<.001$), level 2 ($\beta=-.434, z=-5.3,p<.001$) and level three ($\beta=-.250,z=-4.1, p<.001$).

(H3$_{er}$) Error rates were greater in the crisp conditions than the vague conditions when numerical distance was small, and this difference diminished with increasing numerical distance until it reversed at the biggest numerical distance. % MG NEED TO INSERT STATS FOR THIS INTERACTION -- KvD Can you add this please? If it's a problem, we can omit H3.

So far, these results are in line with the idea of cost reduction. Cost reduction explains the vagueness advantage by claiming that the vague referring expressions place less cognitive load on the comprehender than the crisp referring expressions. It explains the diminishing returns for vagueness in more-discriminable stimuli (i.e., the vagueness by numerical distance interaction) by claiming that load is low in both conditions for the easily-discriminable stimuli, and that therefore there is no extra benefit to be had from vagueness in the easily-discriminable stimuli.

@@

% KvD: I've added some syntactic sugar, in a number of places, to clarify the relation between Ex 1 and Ex 2.

However, the picture painted by these findings might be misleading. First of all, there is a possibly confounding factor. Contrast, from Experiment 1, an expression from the vague condition: `the square with few dots' with an expression from the crisp condition: `the square with 5 dots'. 
One difference is that `few' has the potential for vagueness, whereas `5' is crisp. But
another difference is that `few' is verbal while `5' is numerical, in the sense that a number is mentioned explicitly. 
Since these two differences could not be separated in Experiment 1, the vagueness advantage finding is vulnerable to an alternative interpretation, that what we saw as a vagueness advantage was in contrast an advantage for the verbal form of the quantifier. 
In Experiment 2 we therefore created verbal and numeric versions of each of the vague and crisp instructions so that we could compare vague and crisp conditions while taking account of verbal / numeric format.
%In the present experiment 2 we pitted these alternative interpretations against each other in a factorial design.

Another potential problem with Experiment 1 is the following. Participants chose one of two squares: therefore the `vague' quantifiers (e.g., `few') uniquely identified one square. Recall our definition of vague  -- ``a word is precise if it describes a well-defined set of objects. By contrast, a word is vague if it is not precise''.  In Experiment 1, the quantifiers in the vague conditions did not realise their potential for vagueness. This is because there were no borderline cases of the referent that could make the referent set `not well-defined', and perhaps because using definite articles in the instructions implied that only one option was correct.
Using error feedback in Experiment 1 could have exacerbated this.

To find out what happens when words are used in a context where their potential for vagueness comes to the fore,
Experiment 2 used three arrays so that the vague quantifiers always had more than one possible referent, and used indefinite articles in the vague instructions to avoid the impression that only one response counted as correct, and was carried out without error feedback.
An indication that the potential for vagueness was realised in Experiment 2 is that the borderline response was chosen fairly often: 16\% of the time.

In Experiment 2, an item was a referring expression instruction followed by a set of dot arrays defined by a triple of numbers, representing the number of dots in the left, middle, and right arrays. We used four different triples of numbers: (6,15,24); (16,25,34); (26,35,44); (36,45,54). Each set of arrays had the following properties: it comprised three arrays (instead of two as in Experiment 1); the array representing the central number was always presented in the middle of the three; there were two flanking arrays where one had fewer dots than the central array and the other had more.

Examples of crisp and vague versions of the numerical and verbal instructions follow: the examples assume the array (6,15,24) and reference to the smaller number of dots, such that 6 was classified as the expected response; 15 was classified as the borderline response; and 24 was classified as the incorrect response. In the {\em vague numerical} condition we used \emph{Choose a square with about 10 dots}. None of the squares contained 10 dots. 10 is slightly closer to 6 than to 15, justifying 6 as the best response and 15 as the borderline response. In the {\em vague verbal} condition we used \emph{Choose a square with few dots}. In the {\em crisp numerical} condition we used \emph{Choose the square with 6 dots}, and one square always did contain the number mentioned. For {\em crisp verbal}, we used \emph{Choose the square with the fewest dots}.

% KvD I've re-instated the table below, because this makes it easier to understand the instruction format/selection algorithm confound.

\begin{table}[tbp]
\caption{Table of instructions arranged by condition for the triple (6,15,24)}
\label{instre3}
\begin{tabular}{rll}
&crisp&vague\\
\hline
numerical	&		Choose the square with 6 dots 			& 	Choose a square with about 10 dots			\\
verbal	&		Choose the square with the fewest dots	&	Choose a square with few dots\\
\end{tabular}
\end{table}

\subsection{Hypotheses}
(H1) A main effect RT advantage for vagueness.

(H2) An RT advantage for vagueness both in \emph{numeric} and in \emph{verbal} instructions.

(H3) No large main effect of instruction format (numeric versus verbal). 
%MG If we are right that vagueness was responsible for the vagueness advantage in e 1, that could also have been a number avoidance advantage, then we would not expect that number avoidance, or, instruction format, plays a role (?)

(H4) On the basis of Experiment 1, we would expect faster responses to more discriminable triples (i.e., an effect of triple).

(H5) Participants should make more borderline case choices for vague than crisp instructions.

@@

(H1) The main effect of \emph{vagueness} was to slow responses down, in contrast with Experiment 1, and offering evidence against hypothesis 1 (vague: 2668 ms; crisp: 2450 ms; a difference of 218 ms; $\beta=.06, se=.01, t=4.6, p<.001$).

(H2) Vagueness was disadvantageous in both the \emph{numeric} and \emph{verbal} conditions, offering evidence against hypothesis 2. %MG [? give values of a focussed t test]
The disadvantage for vagueness was greater in the numerical than in the verbal conditions, leading to an interaction effect between vagueness and instruction format ($\beta=-0.13, se= 0.02, t= -6.6, p=0.000$).

(H3) There was a significant effect of \emph{instruction format} with numerical conditions attracting longer responses than the verbal conditions: consistent with Experiment 1 (numeric: 3284 ms; verbal 1866 ms; a difference of 1418 ms; $\beta=.37, se=.07, t=5.1, p<.001$).

(H4) There was a significant main effect of \emph{triple} ($\beta=.06, se=.008, t=7.0, p<.001$) indicating that response times differed in some way according to which triple was presented. 
However further analysis revealed that there was no consistent smooth trend across different triples. 
This effect seems likely to be due to the very fast responses for the smallest triple, which had the largest ratio difference and so may have been particularly discriminable for participants. 
% MG reviews asked for an explanation of the difference for numeric crisp over items. Also reviews mention ratio differences as something that we did not control.

\subsubsection{Borderline responses}
(H5) Participant grand mean percentage of borderline selections was $16.6\%$. A generalized linear mixed model \cite{jaeger2008categorical} was fit to the data for selection of the borderline response, with task, vagueness and item as fixed effects, and with random slopes for task and vagueness and item over participants.  
The distribution of responses over the nearest match square, the borderline square, and the furthest match square are given in Fig, \ref{resultse2}.
Participants were significantly more likely to choose the borderline option for vague instructions than for precise instructions (21.9\% vs 11.3\%, $\beta=.79, se=.25, z=3.2, p<.01$). 
Participants were significantly more likely to choose the borderline square when the instruction used the numerical format rather than the verbal format (30.1\% vs 3.0\%, $\beta=3.57, se=.26, z=13.6, p<.001$).

@@

% KvD I've expanded the following:
However, once again there is a possibly confounding factor. Observe that, in Experiment 2, instruction format (i.e., the choice between numeric and verbal) went hand in hand with might be called the (human) {\bf selection algorithm}: 
 %as shown by the Table of instructions \ref{instructionse2}, 
%the instructions that used a verbal instruction format allowed a comparison algorithm, whereas the instructions that used a numeric format allowed a matching algorithm. 
To see this, consider the task of selecting the dot array that contains ``few dots": to do this, it suffices to {\em compare} the three arrays and select the one that contains the fewest elements. Therefore, our results so far permit the interpretation that what made the instructions in the verbal condition fast is not the fact that they were worded verbally, but that they allowed participants to use a comparison ``algorithm" (which is known to be faster than matching). % KvD I think we need to say more about this, citing some appropriate papers to support this idea.

@@

\begin{table}[htbp]
\centering
\caption{Experiment 3: Instructions, assuming $6,15,24$ dots as the Item} %, and showing \emph{fewer} instead of \emph{more}}
\label{Instructions for e3}
\begin{tabular}{llll}
\hline
vagueness&instruction format&selection task&instruction\\
\hline
crisp 	& numeric		&	matching		& Choose a square with 6 dots \\ 
crisp 	& numeric		&	comparison 	& Choose a square with fewer than 20 dots \\
vague	& numeric		&	matching 		& Choose a square with about 10 dots \\ 
vague 	& numeric		&	comparison 	& Choose a square with far fewer than 20 dots \\ 
\hline
\end{tabular}
\end{table}%
